---
title: "EECA Energy End Use Database analysis"
author: "Ben Anderson (dataknut@icloud.com)"
date: 'Last run at: `r Sys.time()`'
output: 
  bookdown::html_document2:
    fig_caption: yes
    toc: TRUE
    toc_depth: 4
    toc_float: TRUE
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) # but folded by yaml setting

library(dkUtils) # https://github.com/dataknut/dkUtils

myLibs <- c("data.table", # fast data munching
            "ggplot2", # plots obvs
            "plotly"
            )
dkUtils::loadLibraries(myLibs) # load the libs (installs if not present)
```

# Introduction

Analysis of EECA's Energy End Use Data for New Zealand.

 * Data: https://www.eeca.govt.nz/insights/data-tools/energy-end-use-database/
 * Visualisation tool: https://www.eeca.govt.nz/insights/data-tools/energy-end-use-database/

# Data

```{r loadData}
dataFile <- path.expand("~/Dropbox/data/EECA/eeud_data-2024-08-16-657.csv")

dt <- data.table::fread(dataFile)

```

The data gives total energy used (TJ) per year for different purposes by different sectors and by type.

For example, Figure \@ref(fig:totalBySectorGroup) shows total energy use over time by sector group while Figure \@ref(fig:totalByFuel) shows the total by energy (fuel) type.

```{r totalBySectorGroup, fig.cap="Energy use by sector over time"}

make_colPlot <- function(dt, xVar, yVar, fillVar, scaleVar){
  p <- ggplot2::ggplot(dt, aes(x = get(xVar), 
                               y = get(yVar), 
                               fill = get(fillVar))) +
    geom_col() +
    scale_fill_discrete(name = scaleVar)
  return(p)
}

plotDT <- dt[, .(Total_TJ = sum(TJ,na.rm = TRUE)),
             keyby = .(SectorGroup, Period)]

make_colPlot(plotDT, "Period", "Total_TJ", "SectorGroup", "Sector") +
  labs(x = "Year", y = "Total TJ")

message("2022 total")
sum(plotDT[Period == 2022, Total_TJ])

```

```{r totalByFuel, fig.cap="Energy use by fuel group over time"}

plotDT <- dt[, .(Total_TJ = sum(TJ, na.rm = TRUE)),
             keyby = .(FuelGroup, Period)]

make_colPlot(plotDT, "Period", "Total_TJ", "FuelGroup", "FuelGroup") +
  labs(x = "Year", y = "Total TJ")

message("2022 total")
sum(plotDT[Period == 2022, Total_TJ])

```


```{r totalElecBySector, fig.cap="Electricity use by sector over time"}

plotDT <- dt[Fuel %like% "Electricity", .(Total_TJ = sum(TJ, na.rm = TRUE)),
             keyby = .(SectorGroup, Period)]

make_colPlot(plotDT, "Period", "Total_TJ", "SectorGroup", "SectorGroup") +
  labs(x = "Year", y = "Total TJ")
message("2022 total electricity")
sum(plotDT[Period == 2022, Total_TJ])

```

# Residential energy use


```{r descPlot, fig.width=8}
dt[, TJ_pc := 100 * TJ/sum(TJ, na.rm = TRUE), keyby = .(Period)]

res_dt <- dt[Sector %like% "Residential"]

res_dt[, TJ_res_pc := 100 * TJ/sum(TJ), keyby = .(Period)]

ggplot2::ggplot(res_dt,
                aes(x = Period, y = TJ, fill = paste0(EndUseGroup, " - ", Fuel))) +
  geom_col(position = "stack")

ggplot2::ggplot(res_dt,
                aes(x = Period, y = TJ_res_pc, fill = paste0(EndUseGroup, " - ", Fuel))) +
  geom_col(position = "stack")

t <- res_dt[Period == 2022, .(sum_TJ = sum(TJ),
                              TJ_pc = sum(TJ_pc, na.rm = TRUE),
                              TJ_res_pc = sum(TJ_res_pc)), keyby = .(EndUse, Fuel)]

knitr::kable(t, caption = "Summary for 2022", digits = 2)
```

```{r}
res_dt[EndUse %like% "Water", branzCat := "Water heating"]
res_dt[EndUse %like% "Space" , branzCat := "Space heating"]
res_dt[EndUse %like% "Cooking", branzCat := "Cooking"]
res_dt[Fuel %like% "Electricity" & 
         is.na(branzCat), branzCat := "Lighting, refrigeration & other electricity uses"]

bt <- res_dt[Period == 2022 & !is.na(branzCat), .(sum_TJ = sum(TJ),
                                        TJ_pc= sum(TJ_pc, na.rm = TRUE),
                              TJ_res_pc = sum(TJ_res_pc)), keyby = .(branzCat)]
bt[, TJ_branzCat_pc := 100 * sum_TJ/sum(sum_TJ)]
knitr::kable(bt, caption = "Summary for 2022", digits = 2)
```


## Residential electricity use

```{r descStats, fig.width=8}
dt[, TJ_fuel_pc := 100 * TJ/sum(TJ, na.rm = TRUE), keyby = .(Fuel, Period)]

res_elec_dt <- dt[Fuel %like% "Electricity" &
                     Sector %like% "Residential"]

ggplot2::ggplot(res_elec_dt,
                aes(x = Period, y = TJ, fill = EndUseGroup)) +
  geom_col(position = "stack")

```

```{r residentialElecHeat, fig.cap="Electricity used for heat"}
p <- ggplot2::ggplot(res_elec_dt[EndUse %like% "Heat"],
                aes(x = Period, y = TJ, 
                    fill = Technology
                    )) +
  geom_col(position = "stack") +
  theme(legend.position = "bottom")

plotly::ggplotly(p)
```

```{r stats2022}
resElec_2022_pc <- sum(res_elec_dt[Period == 2022, TJ_fuel_pc])
resElec_2022 <- sum(res_elec_dt[Period == 2022, TJ])

resElec_2022_spaceHeat <- sum(res_elec_dt[Period == 2022 &
                                   EndUse %like% "Space Heating", TJ])
resElec_2022_spaceHeat_pc <- sum(res_elec_dt[Period == 2022 &
                                   EndUse %like% "Space Heating", TJ_fuel_pc])

resElec_2022_dt <- res_elec_dt[Period == 2022, .(total_TJ = sum(TJ, na.rm = TRUE),
                                             total_pc = sum(TJ_pc, na.rm = TRUE),
                                             total_elec_pc = sum(TJ_fuel_pc, na.rm = TRUE)),
                           keyby = .(Sector, EndUseGroup, EndUse)]
resElec_2022_dt[, resElec_pc := 100 * total_TJ/sum(total_TJ)]
knitr::kable(resElec_2022_dt, digits = 2, caption = "2022 residential electricity use summary")
```

2022

 * residential = `r prettyNum(resElec_2022, big.mark = ",")` TJ (`r round(resElec_2022_pc,1)` % of all electricity)
 * residential space heating = `r prettyNum(resElec_2022_spaceHeat, big.mark = ",")` TJ (`r round(resElec_2022_spaceHeat_pc,1)` % of all electricity)


```{r residentialHeat}

ggplot2::ggplot(res_elec_dt[EndUse %like% "Space Heat"], aes(x = Period,
                        y = TJ,
                        colour = Technology))+
  geom_line()
  # theme(legend.position = "bottom") +
  # guides(colour=guide_legend(ncol=1))


ggplot2::ggplot(res_elec_dt[EndUse %like% "Space Heat"], aes(x = Period,
                        y = TJ_fuel_pc,
                        colour = Technology))+
  geom_line() +
  labs(y = "% total electricity")

res_elec_dt[, resElec_pc := 100 * TJ/sum(TJ), keyby = .(Period)]

ggplot2::ggplot(res_elec_dt[EndUse %like% "Space Heat"], aes(x = Period,
                        y = resElec_pc,
                        colour = Technology))+
  geom_line() +
  labs(y = "% residential electricity")

```

Oh. Well that's interesting... The % of residential electrity used for these two types of heat have never been updated...
